Stochastic Variance Reduced Policy Gradient

### 2018-02-01 svrpg.tex
changed notation of ideal gradient
made second paragraph of intro more explicit (still missing citations)

%%%CUSTOM COMMANDS%%%
%Math
\newcommand{\realspace}{\mathbb R}      % realspace
\newcommand{\transpose}[1]{{#1}^\texttt{T}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\EV}{\mathbb{E}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator*{\Cov}{\mathbb{C}ov}
\DeclareMathOperator*{\Var}{\mathbb{V}ar}
\newcommand{\EVV}[2][\ppvect \in \ppspace]{\EV_{#1}\left[{#2}\right]}
\newcommand{\norm}[2][\infty]{\left\|#2\right\|_{#1}}
\newcommand{\Dij}[2]{\frac{\partial^{2}{#1}}{\partial{#2}_i\partial{#2}_j}}
\newcommand{\de}{\,\mathrm{d}}
\newcommand{\dotprod}[2]{\left\langle#1,#2\right\rangle}
%RL
\newcommand{\vtheta}{\boldsymbol{\theta}}
\newcommand{\Aspace}{\mathcal{A}}
\newcommand{\Sspace}{\mathcal{S}}
\newcommand{\Tspace}{\mathcal{T}}
\newcommand{\Transition}{\mathcal{P}}
\newcommand{\Reward}{\mathcal{R}}
\newcommand{\stationary}{d_{\rho}^{\pi_{\vtheta}}(s)}
\newcommand{\policy}{\pi_{\vtheta}(a \mid s)}
\newcommand{\pol}{\pi_{\vtheta}}
\newcommand{\trajdistr}{\pi_{\vtheta}(\tau)}
\newcommand{\score}[2]{\nabla\log\pi_{#1}(#2)}
\newcommand{\Qfun}{Q^{\pi_{\vtheta}}(s,a)}
\newcommand{\Vfun}{V^{\pi_{\vtheta}}(s)}
\newcommand{\vTheta}{\boldsymbol{\Theta}}
\newcommand{\vphi}{\boldsymbol{\phi}}
\newcommand{\gradJ}[1]{\nabla J(#1)}
\newcommand{\gradApp}[2]{\hat{\nabla}_{#2}J(#1)}
\newcommand{\eqdef}{\mathrel{\mathop:}=}
\newcommand{\Dataset}{\mathcal{D}}
%Specific
\newcommand{\Ets}[2][t]{\mathbb{E}_{#1\mid s}\left[#2\right]}
\newcommand{\Covts}[3][t]{{\mathbb{C}\text{ov}}_{#1\mid s}\left(#2,#3\right)}
\newcommand{\Varts}[2][t]{{\mathbb{V}\text{ar}}_{#1\mid s}\left[#2\right]}
\newcommand{\gradBlack}[1]{\blacktriangledown J(#1)}
\newcommand{\gradIdeal}[1]{\overline{\blacktriangledown} J(#1)}
\newcommand{\VARRF}{V}
\newcommand{\GRADLOG}{G}
\newcommand{\VARIS}{W}
\newcommand{\HESSLOG}{F}
%%%%%%

@@

\section{Introduction}
%Similarities between RL and SL
On a very general level, machine learning addresses the problem of artificial agents that must select the right actions to solve a given task. The approach of Reinforcement Learning \cite{sutton2011reinforcement} is to learn the best actions by direct interaction with the environment and evaluation of the performance in the form of a reward signal. This makes RL fundamentally different from Supervised Learning (SL), where correct actions are explicitly prescribed by a human teacher (\eg for classification, in the form of class labels). However, the two approaches share many challenges and, as a consequence, solution concepts. The problem of estimating a model from samples, which is at the core of SL, is equally fundamental in RL, whether we choose to model the environment, a value function, or a policy of behavior directly. In particular, when the tasks are characterized by large or continuous state-action spaces, RL needs to borrow the powerful function approximators that are the main subject of study of SL, such as neural networks. Among RL solutions for learning continuous tasks, policy gradient methods bear the closest similarity to SL, since they allow to formulate the RL problem as the unconstrained optimization of an objective function (\ie a negative loss) of policy parameters and data samples.

% SL needs SVRG so RL
In supervised learning, stochastic optimization is often employed for reasons of efficiency. The idea is to evaluate the objective on a random subset of the available data to obtain an approximate gradient direction. This can be much less expensive that taking into account all the data for all parameter updates. However, the sub-sampling of data samples is a source of variance that can potentially compromise convergence, so that speed and variance must be traded off with proper handling of meta-parameters.
Variance-reduced gradient algorithms such as SAG, SVRG and SAGA offer better ways of achieving the optimal trade-off between variance and speed, both in theory and practice. 
In Reinforcement Learning, stochastic optimization is not even a choice, since samples must be sequentially obtained by interacting with an initially unknown environment. In this scenario, containing the variance of the estimates is a problem that cannot be avoided, and methods like SVRG can offer a more efficient alternative

@@

@@

\begin{definition}\label{def:ideal}
The ideal SVRPG gradient estimate is:
\begin{align*}
	\overline{\blacktriangledown}J(\vtheta_t^{s+1}) &\coloneqq 
	\gradJ{\tilde{\vtheta}^s}
	+ \nabla\log\pi(\tau_i \mid \vtheta_t^{s+1})\Reward(\tau_i) 
	- \omega(\tau_i)\nabla\log\pi(\tau_i \mid \tilde{\vtheta}^s)\Reward(\tau_i) \\
	&= \gradBlack{\vtheta_t^{s+1}} - \gradApp{\tilde{\vtheta}^s}{N} + \gradJ{\tilde{\vtheta}^s} \\
	&= \gradBlack{\vtheta_t^{s+1}} - e_s
\end{align*}
\end{definition}

>>>>

%%%CUSTOM COMMANDS%%%
%Math
\newcommand{\realspace}{\mathbb R}      % realspace
\newcommand{\transpose}[1]{{#1}^\texttt{T}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\EV}{\mathbb{E}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator*{\Cov}{\mathbb{C}ov}
\DeclareMathOperator*{\Var}{\mathbb{V}ar}
\newcommand{\EVV}[2][\ppvect \in \ppspace]{\EV_{#1}\left[{#2}\right]}
\newcommand{\norm}[2][\infty]{\left\|#2\right\|_{#1}}
\newcommand{\Dij}[2]{\frac{\partial^{2}{#1}}{\partial{#2}_i\partial{#2}_j}}
\newcommand{\de}{\,\mathrm{d}}
\newcommand{\dotprod}[2]{\left\langle#1,#2\right\rangle}
\newcommand{\dnabla}{\nabla\!\!\!\!\nabla}
%RL
\newcommand{\vtheta}{\boldsymbol{\theta}}
\newcommand{\Aspace}{\mathcal{A}}
\newcommand{\Sspace}{\mathcal{S}}
\newcommand{\Tspace}{\mathcal{T}}
\newcommand{\Transition}{\mathcal{P}}
\newcommand{\Reward}{\mathcal{R}}
\newcommand{\stationary}{d_{\rho}^{\pi_{\vtheta}}(s)}
\newcommand{\policy}{\pi_{\vtheta}(a \mid s)}
\newcommand{\pol}{\pi_{\vtheta}}
\newcommand{\trajdistr}{\pi_{\vtheta}(\tau)}
\newcommand{\score}[2]{\nabla\log\pi_{#1}(#2)}
\newcommand{\Qfun}{Q^{\pi_{\vtheta}}(s,a)}
\newcommand{\Vfun}{V^{\pi_{\vtheta}}(s)}
\newcommand{\vTheta}{\boldsymbol{\Theta}}
\newcommand{\vphi}{\boldsymbol{\phi}}
\newcommand{\gradJ}[1]{\nabla J(#1)}
\newcommand{\gradApp}[2]{\hat{\nabla}_{#2}J(#1)}
\newcommand{\eqdef}{\mathrel{\mathop:}=}
\newcommand{\Dataset}{\mathcal{D}}
%Specific
\newcommand{\Ets}[2][t]{\mathbb{E}_{#1\mid s}\left[#2\right]}
\newcommand{\Covts}[3][t]{{\mathbb{C}\text{ov}}_{#1\mid s}\left(#2,#3\right)}
\newcommand{\Varts}[2][t]{{\mathbb{V}\text{ar}}_{#1\mid s}\left[#2\right]}
\newcommand{\gradBlack}[1]{\blacktriangledown J(#1)}
\newcommand{\gradIdeal}[1]{\dnabla J(#1)}
\newcommand{\VARRF}{V}
\newcommand{\GRADLOG}{G}
\newcommand{\VARIS}{W}
\newcommand{\HESSLOG}{F}
%%%%%%

@@

\section{Introduction}
%Similarities between RL and SL
On a very general level, machine learning addresses the problem of an agent that must select the right actions to solve a given task. The approach of Reinforcement Learning \cite{sutton2011reinforcement} is to learn the best actions by direct interaction with the environment and evaluation of the performance in the form of a reward signal. This makes RL fundamentally different from Supervised Learning (SL), even in its online version, where correct actions are explicitly prescribed by a human teacher (\eg for classification, in the form of class labels). However, the two approaches share many challenges and, as a consequence, solution concepts. The problem of estimating a model from samples, which is at the core of SL, is equally fundamental in RL, whether we choose to model the environment, a value function, or a policy of behavior directly. Moreover, when the tasks are characterized by large or continuous state-action spaces, RL needs to borrow the powerful function approximators that are the main subject of study of SL, such as neural networks.

% SL needs SVRG so RL
In a typical discriminative supervised learning setting, an objective function $J(x\mid\vtheta)$ has to be optimized \wrt to a parameter vector $\vtheta$. The training set $\mathcal{D}_{train}$ is the set of data samples $x$ for which the objective (and its gradient) can be evaluated. In most practical cases, $\mathcal{D}_{train}$ is only a subset of the set $\mathcal{D}$ of all the $x$ of interest, hence we resort to stochastic optimization of the expected value of $J$ over an unknown data distribution. The special case in which $\mathcal{D}_{train} = \mathcal{D}$ is known as Empirical Risk Minimization (ERM), and allows for deterministic optimization. Even in this case, stochastic optimization is often used for reasons of efficiency. The idea is to evaluate the objective on a random subset of the available data to obtain an approximate gradient direction. This can be much less expensive than taking into account all the data for all parameter updates. However, the sub-sampling of data samples is a source of variance that can potentially compromise convergence, so that speed and variance must be traded off with proper handling of meta-parameters.
Variance-reduced gradient algorithms such as SAG, SVRG and SAGA offer better ways of achieving the optimal trade-off between variance and speed, both in theory and practice. Although designed explicitly for ERM, this algorithms address a problem that is equally important, if not more critical, in more general machine learning problems.
In Reinforcement Learning, as in online learning n general, stochastic optimization is rarely a matter of choice, since data must be actively sampled by interacting with an initially unknown environment. In this scenario, containing the variance of the estimates is a problem that cannot be avoided, and methods like SVRG can offer a more efficient alternative, or a complement, to traditional methods such as batch learning.

@@

% Among RL solutions for learning continuous tasks, policy gradient methods bear the closest similarity to SL, since they allow to formulate the RL problem as the unconstrained optimization of an objective function (\ie a negative loss) of policy parameters and data samples.

@@

\begin{definition}\label{def:ideal}
The ideal SVRPG gradient estimate is:
\begin{align*}
	\gradIdeal{\vtheta_t^{s+1}} &\coloneqq 
	\gradJ{\tilde{\vtheta}^s}
	+ \nabla\log\pi(\tau_i \mid \vtheta_t^{s+1})\Reward(\tau_i) 
	- \omega(\tau_i)\nabla\log\pi(\tau_i \mid \tilde{\vtheta}^s)\Reward(\tau_i) \\
	&= \gradBlack{\vtheta_t^{s+1}} - \gradApp{\tilde{\vtheta}^s}{N} + \gradJ{\tilde{\vtheta}^s} \\
	&= \gradBlack{\vtheta_t^{s+1}} - e_s
\end{align*}
\end{definition}
